---
title:  |
  <center> Evaluating a hypothetical land use decision on   </center>
  <center> land consolidation for smallholder farmers </center>
  <center>  in Kenya </center>
  <br>
  <center><i> supplementary material 1</i></center>
  <br>
author: "Hannah Kamau, Cory Whitney"
bibliography:
- bib/references.bib
- bib/packages.bib
link-citations: true
output:
  html_document:
     toc: true
---

```{r setup, include=FALSE}

library(decisionSupport)
library(tidyverse)
library(magrittr)
library(patchwork)   # for the `+`, `/`, and wrap_plots() figure composition
library(kableExtra)
library(DT)
library(grid)
library(scales)

set.seed(254)

# --- shared figure style ---------------------------------------------------
theme_lc <- theme_bw(base_size= 12) + 
  theme(axis.title  = element_text(face = "bold"),
        legend.title = element_blank())

pal_decision <- c("Land consolidation" = "#1a80bb",  # blue
                  "Maintain farms"      = "#d1495b")  # muted crimson 
```

# Introduction

This document is a supplementary resource describing the land consolidation model
used to simulate the outcomes of the decision to consolidate land in Kenya by
smallholder farmers. The data are expert estimates collected in March 2024, after
co-creating a conceptual model with stakeholders, supplemented with published
literature where available and were unavailable other stakeholders not in in the workshop were contacted to discuss the estimates. 

# Required packages

The core R packages are `decisionSupport` [@R-decisionSupport], `tidyverse` [@R-tidyverse], and `ggplot2` [@ggplot22016].

```{r write_package_bib, include=FALSE}
# Write package citations to a .bib file for the reference list.
knitr::write_bib(
  c(.packages(), "decisionSupport", "ggplot2", "tidyverse",
    "magrittr", "patchwork", "kableExtra", "DT", "grid", "scales"),
  "bib/packages.bib")
```

# From conceptual model to a land consolidation function

The conceptual model, co-created with stakeholders, reflects the costs, benefits,
and risks that could arise when shifting from small, highly diversified farms to a
model where farmers consolidate their land under external management.

**Costs a farmer could incur under land consolidation**

* Planning costs (legal fees, knowledge acquisition, stationery, printing)
* Loss of natural assets, e.g. perennial plants such as trees
* Food costs no longer saved
* Household upkeep cost prior to the first payment
* Medical bills due to a reduced active lifestyle

**Benefits a farmer could gain under land consolidation**

* Lease income
* Saved production costs
* Alternative income from on- or off-farm activities
* Better childhood outcomes
* Social cohesion
* Saved medical bills related to land-use accidents/diseases

**Costs incurred by a farmer who remains in the current system ("do nothing")**

* Planning costs (incurred whether or not the farmer changes)
* Medical bills related to land-use accidents/diseases and stress
* Production costs

**Benefits under the "do nothing" option**

* Land security (valued via trees)
* Annual crop yield
* Saved food costs

Both options face risks: farmers who shift face social risks from freed-up time,
while those who do nothing face crop-related risks.

*Shifting to land consolidation is modeled over a 25-year period, during which the
farmer can no longer farm the consolidated parcel but receives lease income
commensurate with its size.*

The `land_consolidation_function` defines these costs, benefits, and risks for both options. The **decision** is the difference between the two options.

# Inputs, model function, and test variables

The input file which contains the variables and estimates --> `land_consolidation_input.csv`.
Market scenario file based on value of information --> `market_scenario.csv` which is used later. 

```{r load_inputs}
input_path    <- "land_consolidation_input.csv"
base_estimate <- decisionSupport::estimate_read_csv(input_path, sep = ";")
input_table   <- read.csv(input_path, sep = ";")
```

```{r make_variables}

make_variables <- function(est, n = 1) {
  x <- decisionSupport::random(rho = est, n = n)
  for (i in colnames(x)) assign(i, as.numeric(x[1, i]), envir = .GlobalEnv)
}
make_variables(base_estimate)
```

```{r source_model_function}
source("land_consolidation_function.R")
```

# Monte Carlo simulation

We run a Monte Carlo simulation of the model function using the inputs in
`land_consolidation_input.csv`, and in this model we use 10,000 model runs. Full model described here can be found in the `land_consolidation_function.R`

```{r run_mc_simulation}
mcSimulation_results <- decisionSupport::mcSimulation(
  estimate          = base_estimate,
  model_function    = land_consolidation_function,
  numberOfModelRuns = 1e4,
  functionSyntax    = "plainNames")
```

# Results

## Distribution of the outcomes
We present 2 key results:- 
(i) The distribution of the 2 decision options (i) land consolidation and (ii) maintain farms; and (ii) the distribution of decision do which is reflected by the difference of the options. 

```{r fig1_outcome_distributions, message= FALSE, fig.cap="*Distribution of the outcomes of land consolidation vs do-nothing options given as net present value per hectare. Based on 10,000 Monte Carlo model runs.*"}

# Assemble model inputs (x) and outputs (y), then express NPVs per hectare.
mc_output <- cbind(as.data.frame(mcSimulation_results$y),
                   as.data.frame(mcSimulation_results$x)) %>%
  mutate(NPV_per_ha          = benefit_lc_kes        / ha_per_hh,
         No_NPV_per_ha       = benefit_status_quo_kes / ha_per_hh,
         NPV_per_ha_decision = net_benefit_lc_kes     / ha_per_hh)

# Summary statistics (million KES).
summarised_stats <- mc_output %>%
  select(NPV_per_ha, No_NPV_per_ha, NPV_per_ha_decision) %>%
  summarise(across(everything(), list(
    min         = ~min(., na.rm = TRUE) / 1e6,
    q05         = ~quantile(., 0.05, na.rm = TRUE) / 1e6,
    median      = ~median(., na.rm = TRUE) / 1e6,
    q95         = ~quantile(., 0.95, na.rm = TRUE) / 1e6,
    max         = ~max(., na.rm = TRUE) / 1e6,
    pc_positive = ~mean(. > 0, na.rm = TRUE) * 100))) %>%
  pivot_longer(everything(),
               names_to = c("variable", "statistic"),
               names_pattern = "^(.*)_(.*)$", values_to = "value") %>%
  pivot_wider(names_from = statistic, values_from = value) %>%
  mutate(across(min:max, \(x) round(x, 2)))
summarised_stats

# (a) Overlaid densities of the two options (trimmed to the 5-95% range).
mc_output %>%
  select(NPV_per_ha, No_NPV_per_ha) %>%
  pivot_longer(everything(), names_to = "variables", values_to = "project_outcome") %>%
  group_by(variables) %>%
  filter(project_outcome >= quantile(project_outcome, 0.05, na.rm = TRUE),
         project_outcome <= quantile(project_outcome, 0.95, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(variables = factor(variables, levels = c("NPV_per_ha", "No_NPV_per_ha"),
                            labels = c("Land consolidation", "Maintain farms"))) %>%
  ggplot(aes(project_outcome / 1e6, colour = variables, fill = variables)) +
  geom_density(alpha = 0.3) +
  scale_colour_manual(values = pal_decision) +
  scale_fill_manual(values = pal_decision) +
  labs(x = "NPV per hectare (million KES)", y = "Density", title = "") + #Overall outcomes
  theme_lc +
  theme(legend.position = "top")
```

```{r fig2_decision_distributions, message= FALSE,  fig.cap="*Distribution of the decision outcomes given as net present value per hectare. Based on 10,000 Monte Carlo model runs.*" }

# (b) The decision (net benefit), coloured by sign, with directional guidance.
percentages <- mc_output %>%
  summarise(negative = mean(net_benefit_lc_kes < 0,  na.rm = TRUE) * 100,
            positive = mean(net_benefit_lc_kes >= 0, na.rm = TRUE) * 100)
negative_percent <- round(percentages$negative, 2)
positive_percent <- round(percentages$positive, 2)

 mc_output %>%
  mutate(outcome_category = ifelse(net_benefit_lc_kes >= 0, "positive", "negative")) %>%
  ggplot(aes(net_benefit_lc_kes / 1e6, fill = outcome_category)) +
  geom_histogram(breaks = seq(-90, 90, by = 1)) +
  scale_fill_manual(values = c(negative = "#1a80bb", positive = "#d1495b")) +
  guides(fill = "none") +
  annotate("text", label = paste0(negative_percent, " %"), x = -34, y = 200) +
  annotate("text", label = paste0(positive_percent, " %"), x =  20, y = 200) +
  labs(x = "NPV per hectare (million KES)", y = "Frequency",
       title = "") + # Decision outcome for land consolidation
  theme_lc +
  theme(plot.margin = margin(0, 0, 90, 0)) +
  coord_cartesian(clip = "off") +
  annotation_custom(linesGrob(arrow = arrow(length = unit(0.25, "cm"), type = "open"),
                              gp = gpar(col = "black")),
                    xmin = 5, xmax = 45, ymin = -75, ymax = -75) +
  annotation_custom(textGrob("'Consolidate land'\npreferable", x = unit(0.8, "npc"),
                             y = unit(-0.1, "npc"), just = "left", gp = gpar(fontsize = 10)),
                    xmin = 50, xmax = 50, ymin = -75, ymax = -75) +
  annotation_custom(linesGrob(x = unit(c(1, 0), "npc"),
                              arrow = arrow(length = unit(0.25, "cm"), type = "open"),
                              gp = gpar(col = "black")),
                    xmin = -5, xmax = -45, ymin = -75, ymax = -75) +
  annotation_custom(textGrob("'Maintain farms'\npreferable", x = unit(0.8, "npc"),
                             y = unit(-0.1, "npc"), just = "left", gp = gpar(fontsize = 10)),
                    xmin = -85, xmax = -85, ymin = -75, ymax = -75)

```

## Variable importance and Expected Value of Perfect Information

We use variable in projection (VIP) to show important variables in the model and value of information is visualized using the expected value of perfect information (EVPI).  

```{r fig3_vip, message=FALSE, fig.cap="*Sensitivity using important variables from PLS regression (VIP scores). Based on 10,000 Monte Carlo model runs.*"}

pls_result <- plsr.mcSimulation(object = mcSimulation_results,
                                resultName = names(mcSimulation_results$y[3]), ncomp = 1)
decisionSupport::plot_pls(
  pls_result, input_table = input_table, threshold = 1,
  pos_color = "#1a80bb", neg_color = "#ea801c",  x_axis_name = "VIP score",) +
  scale_y_discrete(labels = \(y) str_wrap(y, width = 15)) +
  labs(title = " ") + #Variable importance
  theme_lc + theme(legend.position = "bottom", 
                   legend.title = element_blank())
```

```{r fig4_evpi, message=FALSE, fig.cap="*Value-of-information analyses showing key uncertainties from the EVPI metric. Based on 10,000 Monte Carlo model runs.*"}

mcSimulation_table <- data.frame(mcSimulation_results$x, mcSimulation_results$y[1:3])
if (!exists("evpi")) {
  evpi <- decisionSupport::multi_EVPI(mc = mcSimulation_table, first_out_var = "benefit_lc_kes")
}

plot_evpi(evpi, decision_vars = "net_benefit_lc_kes",
                   input_table = input_table, bar_color = "#1a80bb",
                   x_axis_name = "EVPI in million KES",
                   new_names = "") +
  scale_y_discrete(labels = \(y) str_wrap(y, width = 20)) +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6)) +
  labs(title = "") + #Value of information
  theme_lc

```


# Monte Carlo simulation with scenarios

Based on two most critical variables in Figure 3 — maize and lease prices — we built scenarios showing how the decision changes as those prices vary. 

Requirements: price bounds for both maize and lease prices. 

For maize-price bounds we obtained data from daily submissions to the Ministry of Agriculture (https://amis.co.ke/site/market). We used national level data and for Murang'a County that was available (2005–2024).

```{r figS1_maize_prices, warning=FALSE,  fig.cap="***Figure S1.*** *Distribution of dry white maize prices in Kenya (a) and in Murang'a County (b).*"}

# Read every maize-price CSV from the folder and clean the price columns.
maize_files <- list.files("maize_prices/", pattern = "\\.csv$", full.names = TRUE)

combined_df <- maize_files %>%
  map(\(f) read.csv(f, sep = ";") %>% mutate(Supply.Volume = as.numeric(Supply.Volume))) %>%
  list_rbind() %>%
  mutate(across(c(Wholesale, Retail),
                \(v) as.numeric(na_if(trimws(gsub("/Kg|,", "", v)), "-"))),
         Date = as.Date(Date))

price_boxplot <- function(df, title) {
  df %>%
    select(Wholesale, Retail) %>%
    pivot_longer(everything(), names_to = "market_type", values_to = "prices") %>%
    ggplot(aes(factor(market_type), prices)) +
    geom_boxplot(width = 0.1, colour = "forestgreen") +
    labs(x = "Market type", y = "Price (KES/kg)", title = title) +
    theme_minimal(base_size = 12) +
    theme(axis.title = element_text(face = "bold"))
}

 price_boxplot(combined_df, "a) Dry maize prices, national") +
         price_boxplot(filter(combined_df, County == "Muranga"), "b) Dry maize prices, Murang'a")

```
Maize prices have varied over time.

```{r figS2_maize_price_timeseries, warning=FALSE,  fig.cap="***Figure S2.*** *Monthly mean maize prices (KES/kg) across the recorded markets: (a) the full series and (b) 2021–2024. Retail and wholesale prices are shown separately.*"}

price_timeseries <- function(df, title) {
  df %>%
    mutate(month = as.Date(format(Date, "%Y-%m-01"))) %>%
    group_by(month) %>%
    summarise(Retail    = mean(Retail,    na.rm = TRUE),
              Wholesale = mean(Wholesale, na.rm = TRUE),
              .groups   = "drop") %>%
    pivot_longer(c(Retail, Wholesale), names_to = "Price type", values_to = "price") %>%
    ggplot(aes(month, price, colour = `Price type`)) +
    geom_line(linewidth = 1) +
    labs(x = "Year", y = "Mean price (KES/kg)", title = title, colour = NULL) +
    theme_minimal(base_size = 12) +
    theme(axis.title  = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, hjust = 1))
}

 price_timeseries(combined_df, "a) Monthly maize prices over time") +
         price_timeseries(
           filter(combined_df, Date >= as.Date("2021-01-01"),
                               Date <= as.Date("2024-12-31")),
           "b) Monthly maize prices, 2021–2024") +
         plot_layout(guides = "collect") &
         theme(legend.position = "bottom")
```

The `decisionSupport::scenario_mc` function updates specific variables from a scenario file. Our `market_scenarios.csv` updates maize and lease prices.

***Table 1.*** *Model input variables updated to generate the market scenarios, chosen from the EVPI results (Figure 2).*

```{r scenario_inputs_table}
scenarios  <- read.csv("market_scenarios.csv", sep = ";")
if (knitr::is_latex_output()) { 
  scenarios %>%
    kbl(booktabs = TRUE, row.names = FALSE) %>%
    kable_styling(latex_options = c("striped", "scale_down"))
} else {
  scenarios %>%
    datatable(extensions = "Buttons", rownames = FALSE,
              options = list(dom = "Blfrtip", buttons = c("copy", "csv", "excel"),
                             scrollX = TRUE, autoWidth = TRUE))
}
```

```{r run_scenario_mc_all}

land_consolidation_scenarios <- scenario_mc(
  base_estimate     = base_estimate,
  scenarios         = scenarios,
  model_function    = land_consolidation_function,
  numberOfModelRuns = 1e4,
  functionSyntax    = "plainNames")
```

```{r scenario_helpers, include=FALSE}

run_scenario <- function(col, n = 1000) {
  scenario_mc(base_estimate = base_estimate,
              scenarios = scenarios[, c(1, 2, col)],
              model_function = land_consolidation_function,
              numberOfModelRuns = n, functionSyntax = "plainNames")
}

plot_decision_density <- function(result, title) {
  ggplot(result$y, aes(net_benefit_lc_kes / 1e6)) +
    geom_histogram(aes(y = after_stat(density)), colour = "black", fill = "white", bins = 30) +
    geom_density(alpha = 0.3, fill = "#FFD92F") +
    labs(x = "Decision (million KES)", y = "Relative probability", title = title) +
    theme_bw(base_size = 11)
}
```

# Impact of the market-price scenarios

Table 2 compares the net benefit (the decision) of each scenario against the baseline.

```{r table2_scenario_summary}
# Standardise the combined scenario run to million KES per hectare.
land_consolidation_scenarios$y$net_benefit_lc_per_ha_MKes <-
  land_consolidation_scenarios$y$net_benefit_lc_kes /
  land_consolidation_scenarios$x$ha_per_hh / 1e6

mcSimulation_results$y$net_benefit_lc_per_ha_MKes <-
  mcSimulation_results$y$net_benefit_lc_kes /
  mcSimulation_results$x$ha_per_hh / 1e6
baseline_lc <- mcSimulation_results$y$net_benefit_lc_per_ha_MKes

# Pull each scenario's per-ha net benefit from the single combined run.
scen_keys <- c("low_mp", "med_mp", "high_mp", "low_lp", "med_lp", "high_lp",
               "low_mp_lp", "low_mp_med_lp", "low_mp_high_lp", "med_mp_low_lp",
               "med_mp_lp", "med_mp_high_lp", "high_mp_low_lp", "high_mp_med_lp", "high_mp_lp")
scen_values <- setNames(lapply(seq_along(scen_keys), function(i) {
  land_consolidation_scenarios$y$net_benefit_lc_per_ha_MKes[
    land_consolidation_scenarios$x$Scenario == paste0("Scen_", i)]
}), scen_keys)
list2env(scen_values, envir = environment())

# Display labels, shared by the table and figures.
scenario_labels <- c(
  baseline_lc = "Baseline",
  low_mp = "Low MP", med_mp = "Medium MP", high_mp = "High MP",
  low_lp = "Low LP", med_lp = "Medium LP", high_lp = "High LP",
  low_mp_lp = "Low MP and LP", low_mp_med_lp = "Low MP and Medium LP",
  low_mp_high_lp = "Low MP and High LP", med_mp_low_lp = "Medium MP and Low LP",
  med_mp_lp = "Medium MP and LP", med_mp_high_lp = "Medium MP and High LP",
  high_mp_low_lp = "High MP and Low LP", high_mp_med_lp = "High MP and Medium LP",
  high_mp_lp = "High MP and LP")

qs <- c(0, 0.05, 0.5, 0.95, 1)
all_series <- c(list(baseline_lc = baseline_lc), scen_values)
scenarios_summary_df <- map_dfr(all_series, \(v) round(quantile(v, qs, na.rm = TRUE), 2)) %>%
  as.data.frame()
colnames(scenarios_summary_df) <- c("Minimum", "5% quantile", "Median", "95% quantile", "Maximum")
scenarios_summary_df <- tibble(`Market price scenario` = scenario_labels[names(all_series)],
                               scenarios_summary_df)

if (knitr::is_latex_output()) {
  kable(scenarios_summary_df, format = "latex", booktabs = TRUE, longtable = TRUE, 
        caption = "Table 2. Summary statistics of the net benefit (decision) by market-price scenario, in million KES per hectare. Based on 10,000 runs.", align = c("l", rep("c", 5))) %>%
    kable_styling(latex_options = c("striped", "repeat_header", "scale_down"), font_size = 8) %>%
    add_header_above(c("Market price scenario" = 1, "Quantiles" = 5), underline = TRUE)
} else {
  kable(scenarios_summary_df, format = "html", caption = "<i><b>Table 2.</b> Summary statistics of the net benefit (decision) by market-price scenario, in million KES per hectare. Based on 10,000 runs.</i>", align = c("l", rep("c", 5))) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
    add_header_above(c("Market price scenario" = 1, "Quantiles" = 5), underline = TRUE)
}
```

Preference is the likelihood of a positive net benefit (decision) in each scenario.
We also examine how each scenario shifts the net benefit relative to the baseline.

```{r fig5_scenario_change, fig.cap="*Change in the decision (scenario minus baseline), per hectare. Point = median change; bar = 90% interval. Colour shows whether the scenario makes consolidation more or less attractive. Based on 10,000 runs.*"}

scenario_order <- c("low_mp", "med_mp", "high_mp",
                     "low_lp", "med_lp", "high_lp",
                     "low_mp_lp", "low_mp_med_lp", "low_mp_high_lp",
                     "med_mp_low_lp", "med_mp_lp", "med_mp_high_lp",
                     "high_mp_low_lp", "high_mp_med_lp", "high_mp_lp")

scenario_labels <- c(
  low_mp          = "Low maize price",
  med_mp          = "Medium maize price",
  high_mp         = "High maize price",
  low_lp          = "Low lease price",
  med_lp          = "Medium lease price",
  high_lp         = "High lease price",
  low_mp_lp       = "Low maize price and lease price",
  low_mp_med_lp   = "Low maize price and medium lease price",
  low_mp_high_lp  = "Low maize price and high lease price",
  med_mp_low_lp   = "Medium maize price and low lease price",
  med_mp_lp       = "Medium maize price and lease price",
  med_mp_high_lp  = "Medium maize price and high lease price",
  high_mp_low_lp  = "High maize price and low lease price",
  high_mp_med_lp  = "High maize price and medium lease price",
  high_mp_lp      = "High maize price and lease price")

change_summary <- as_tibble(scen_values) %>%
  mutate(across(everything(), \(v) v - baseline_lc)) %>%   # per-run change vs baseline
  pivot_longer(everything(), names_to = "scenario", values_to = "change") %>%
  group_by(scenario) %>%
  summarise(p05    = quantile(change, 0.05, na.rm = TRUE),
            median = median(change, na.rm = TRUE),
            p95    = quantile(change, 0.95, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(label     = scenario_labels[scenario],
         label     = factor(label, levels = rev(scenario_labels[scenario_order])),
         direction = ifelse(median >= 0, "More attractive", "Less attractive"))

 ggplot(change_summary,
               aes(x = median, y = label, colour = direction)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_pointrange(aes(xmin = p05, xmax = p95), linewidth = 0.7, fatten = 3) +
  scale_colour_manual(values = c("More attractive" = "#1a80bb",
                                 "Less attractive" = "#d1495b"), name = NULL) +
  labs(x = "Change in NPV per hectare vs baseline (million KES)",
       y = NULL) + #,
       # title = "How each market-price scenario shifts the decision for land consolidation") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        axis.title = element_text(face = "bold"))

```

## References
